Continuous developmental change can explain discontinuities in word learning

### 2019-02-01 cogsci/cogsci.tex
final version cogsci

Nevertheless, at least in some cases, development may only appear to be
stage-like. This appearance can be due, for example, to the use of a
cognitively-demanding task which may mask learning, or to the use of
statistical thresholding (in particular, p-value \textless{} 0.05) which
can create a spurious dichotomy between success and failure in observing
a given behavior. In such cases, positing discontinuous stages is
unnecessary. Instead, a continuous model--involving similar
representations across the lifespan--may provide a simpler and more
transparent account of development.

We use a case study from word learning literature. Stager \& Werker
(1997) first showed that children's early ability to distinguish similar
sounds does not automatically translate into word learning skills.
Indeed, though infants around 14-month old can distinguish similar sound
pairs such as ``dih'' and ``bih'', they appear to fail in mapping this
pair to two different objects. Subsequent research has focused on
proposing possible explanations for this observed gap between speech
perception and word learning (e.g., Fennell \& Waxman, 2010; Hofer \&
Levy, 2017; Rost \& McMurray, 2009; Stager \& Werker, 1997). In this
work we focus, rather, on the mechanism of development.

@@

Subsequent findings have suggested otherwise. On the one hand,
14-month-olds--who typically fail in the original task--succeed when an
easier testing method is used, even under the same learning conditions
(Yoshida, Fennell, Swingley, \& Werker, 2009). They also succeed when
uncertainty is mitigated via disambiguating cues (e.g., Thiessen, 2007).
On the other hand, adults show patterns of learning similar to those
shown by 14-month-olds when the task is more challenging and when
similarity between words increases (Pajak, Creel, \& Levy, 2016; White,
Yee, Blumstein, \& Morgan, 2013).

@@

\begin{SCfigure}
\centering
\includegraphics[width=1.5in]{figs/model.png}
\caption{Graphical representation of our model. Circles indicate random variables (shading indicates observed variables). The squares indicates fixed model parameters.}
\label{fig:model}
\end{SCfigure}

A similar probabilistic structure was used by Lewis \& Frank (2013) to
model concept learning, and by Hofer \& Levy (2017) to model spoken word
learning. However, the first study assumed that the sounds are heard
unambiguously, and the second assumed the concepts are observed
unambiguously. In our model, we assume that both labels and concepts are
observed with a certain amount of perceptual noise, which we assume, for
simplicity, that it is captured by a normal distribution:

@@

The learner hears a sound \(s\) and has to decide which object \(o\)
provides an optimal match to this sound. To this end, the learner has to
compute the probability \(P(o|s)\) for all possible objects. This
probability can be computed by summing over all possible concepts and
labels:
\[P(o|s)=\sum_{C,L} P(o, C, L| s) \propto \sum_{C,L} P(o, C, L, s) \]
The joint probability \(P(o, C, L, s)\) is obtained by factoring the
Bayesian network in Figure \ref{fig:model}:
\[P(o,C,L,s) = P(s|L)P(L|C)P(C|o)P(o) \] which could be transformed
using Bayes rule into:

\[P(o,C,L,s) = P(s|L)P(L|C)P(o|C)P(C) \] Finally, assuming that the
concepts' prior probability is uniformly
distributed\footnote{This is a reasonable assumption given the similarity of the concepts used in each naming situation. See the stimuli sub-section in Experiment.},
we obtain the following expression, where all conditional dependencies
are now well defined:

@@

We use the model to predict performance in the word learning task
introduced by Stager \& Werker (1997), and a testing method similar to
the one used by Yoshida et al. (2009). In this task, participants are
first exposed to the association between pairs of nonsense words (e.g.,
``lif''/``neem'') and pairs of objects. The word-object associations are
introduced sequentially. After this exposure phase, participants perform
a series of test trials. In each of these trials, one of the two sounds
is uttered (e.g., ``lif'') and participants choose the corresponding
object from the two alternatives. An overview of the task is shown in
Figure \ref{fig:task}.

We use the general Equation 1 to derive the exact analytical expression
for the probability of accurate responses \(p(o_T | s)\) (target object
\(o_T\) given a sound \(s\)) in the simple case of two-alternative
forced choice in the testing phase of our experimental task:

@@

In this experiment, we tested participants in the word learning task
introduced above (Figure \ref{fig:task}). More precisely, we explored
the predictions related to both distinctiveness and precision. Sound
similarity (\(\Delta s\)) and object similarity (\(\Delta o\)) were
varied simultaneously in a within-subject design. Two age groups
(preschool children and adults) were tested on the same task to explore
whether development can be characterized with the uncertainty
parameters, \(\sigma_C\) and \(\sigma_L\). The Experiment, sample size,
exclusion criteria and the model's main predictions were pre-registered.

@@

We planned to recruit a sample of N=60 children ages 4-5 years from the
Bing Nursery School on Stanford University's campus. So far, we have
collected data from n=47 children. An additional n=28 children
participated but were removed from analyses because they were not above
chance on the catch trials. We also planned to recruit a sample of N=60
adults on Amazon Mechanical Turk. This study used sata from n=50 adult
participants. An additional n=10 participants were tested but excluded
because of low scores on the catch trials (n=9 ) or because they were
familiar with the non-English sound stimuli we used in the adult
experiment (n=1).

@@

In a separate survey, \(N=20\) participants recruited on Amazon
Mechanical Turk evaluated the similarity of each sound and object pair
on a 7-point scale. We scaled responses within the range {[}0,1{]}. Data
are shown in Figure \ref{fig:stim}, for each stimuli group. This data
will be used in the models as the perceptual distance of sound pairs
(\(\Delta s\)) and object pairs (\(\Delta o\)).

@@

Each age group saw only two of the three levels of similarity described
in the previous sub-section: ``different'' vs. ``intermediate'' for
preschoolers, and ``intermediate'' vs. ``similar'' for adults. We made
this choice in the light of pilot studies showing that adults were at
ceiling with ``different'' sounds/objects, and children were at chance
with the ``similar'' sounds/objects. That said, this difference in the
level of similarity is accounted for in the model through using the
appropriate perceptual distance used in each age group (Figure
\ref{fig:stim}).

@@

\caption{\label{fig:stim}Distances for both sound and object pairs from an adult norming study. Error bars represent 95\% confidence intervals.}\label{fig:stim}
\end{figure}
\end{CodeChunk}

The experiment was a 2x2 within-subjects factorial design with four
conditions: high/low sound similarity crossed with high/low visual
object similarity. Besides the 4 conditions, we also tested participants
on a fifth catch condition which was similar in its structure to the
other ones, but was used only to select participants who were able to
follow the instructions and show minimal learning.

@@

Preschoolers were tested at the nursery school using a tablet, whereas
adults used their own computers to complete the same experiment online.
Participants were tested in a sequence of five conditions: the four
experimental conditions plus the catch condition. In each condition,
participants saw a first block of four exposure trials followed by four
testing trials, and a second block of two exposure trials (for memory
refreshment) followed by an additional four testing trials.

@@

\caption{\label{fig:data_all}Accuracy of novel word recognition as as a function of the sound distance, the object distance, and the age group (preschool children vs. Adults). We show both experimental results (top) and model predictions (bottom). Error bars represent 95\% confidence intervals.}\label{fig:all_data}
\end{figure*}
\end{CodeChunk}

@@

Experimental results are shown in Figure 6 (top). We first analyzed the
results using a mixed-effects logistic regression with sound distance,
object distance and age group as fixed effects, and with a maximal
random effects structure (allowing us to take into account the full
nested structure of our data) (Barr, Levy, Scheepers, \& Tily, 2013). We
found main effects for all the fixed effects in the regression, i.e.,
the sound distance (\(\beta =\) 0.44, \(p\) \textless{}
0.001)--replicating previous findings, the object distance (\(\beta =\)
0.5, \(p\) \textless{} 0.001)--confirming the new prediction of our
model, and age group (\(\beta =\) 0.43, \(p\) \textless{}
0.001)--showing that performance improves with age. Besides, we found
two-way interactions between sound distance and age (\(\beta =\) 0.17,
\(p\) \textless{}0.05) and between object distance and age (\(\beta =\)
0.18, \(p\) \textless{}0.05).

Second, we fit our model (using Equation 2) to the participants'
responses in each age group. The values of \(\Delta s\) and \(\Delta o\)
were set based on data from the similarity judgment task (Figure
\ref{fig:stim}). Thus, the model has two degrees of freedom for each
group, i.e., \(\sigma_C\) and \(\sigma_L\). Figure \ref{fig:data_all}
(bottom) shows the predictions. The model captures the qualitative
patterns in both age groups: starting from a low accuracy recognition
when both the sound and object distances are small, the model correctly
predicts an increase in accuracy when either the sound distance or the
object distance increases. Besides, accuracy is correctly predicted to
be maximal when both the sound and object distances are high.

The values of the parameters were as follows. Children had a
label-specific uncertainty of \(\sigma_S =\) 0.9 {[}0.68,
1.11{]}\footnote{All uncertainty intervals in this paper represent 95\% Confidence Intervals.},
and a concept-specific uncertainty of \(\sigma_C =\) 0.29 {[}0.1,
0.49{]}. Adults had a label-specific uncertainty of \(\sigma_S =\) 0.14
{[}-0.02, 0.3{]}, and a concept-specific uncertainty of \(\sigma_C =\)
0.19 {[}0.04, 0.34{]}. As predicted, the uncertainty parameters were
larger for children than they were for adults, showing that the
probabilsitic representations becomes more refined (that is, \(\sigma\)
becomes smaller) across development.

The models explained the majority of the variance in the participants'
mean responses (\(R^2=\) 0.86, for the combined adult and children's
data). To investigate whether the model's predictive power is due to
overfitting, we fit a simplified version with only one degree of freedom
(i.e., a single variance common to both sounds and objects). This
single-variance model explained as much variance in the mean responses
(\(R^2=\) 0.87). It also captured the main qualitative patterns (graph
not shown), suggesting that the explanatory power of the model is
largely due to its structure, rather than its degrees of freedom.

An unexpected outcome was that adult participants deviated slightly from
the model's numbers: While the model predicted accuracy to be more
sensitive to object distance when sound distance is higher (and
vice-versa), adult participants showed the opposite pattern. This
deviation is interesting as it suggests that participants are more
likely to pay attention and integrate additional sources of information
when ambiguity is higher.
\footnote{It is unclear whether the same pattern is found in children as the difference between accuracy scores is generally small compared to the uncertainty of the measurements.}

@@

This paper explored the idea that some seemingly stage-like patterns in
cognitive development can be characterized in a continuous fashion. We
used as a case study the seminal work of Stager \& Werker (1997) showing
a discrepancy between children's speech perception abilities and their
word learning skills. While most previous work debated the source of
this discrepancy, here we were more interested in how it reflects
developmental change.

@@

The model made a new prediction to which we provided the first
experimental test: Learning similar words is not only modulated by the
similarity of their phonological forms, but also by the visual
similarity of their semantic referents. More generally, since visual
similarity is an early organizing feature in the semantic domain (e.g.,
Wojcik \& Saffran, 2013), our finding suggests that children may
prioritize the acquisition of words that are quite distant in the
semantic space. This suggestion is supported by recent findings based on
the investigation of early vocabulary growth (Engelthaler \& Hills,
2017; Sizemore, Karuza, Giusti, \& Bassett, 2018).

One limitation of this work is that the model was fit to data from
children at a relatively older age (4-5 years old) than what is
typically studied in the literature (around 14 month-old). Nonetheless,
this choice was related to two measurement difficulties. First, model
fitting requires several, reliably-measured data points. In our case,
data collection involved presenting participants with several trials
across four conditions in a between-subject design. It would have been
challenging to obtain these measures with infants. Second, the study of
infants requires using specific testing methods. This fact would have
made the quantitative study of development challenging. Indeed, it is
not clear how data from these methods would directly compare to data
collected using more typical methods in older children and adults.

In sum, this paper proposes a model that accounts for the development of
an important aspect of word learning. The findings show that data can be
explained based on a continuous process operating over similar
representations across development. We used a case from word learning as
an example, but the same idea might apply to other aspects of cognitive
development that are typically thought of as stage-like (e.g.,
acquisition of a theory of mind). Computational models such as the one
proposed here can help us investigate the extent to which such
discontinuities emerge due to genuine qualitative changes, and the
extent to which they reflect the granularity of the researchers' own
measurement tools.

>>>>

Nevertheless, at least in some cases, development may only appear to be
stage-like. This appearance can be due, for example, to the use of a
cognitively-demanding task which may mask learning, or to the use of
statistical thresholding (in particular, p-value \textless{} 0.05) which
can create a spurious dichotomy between success and failure in observing
a given behavior. In such cases, positing discontinuous stages is
unnecessary. Instead, a continuous model---involving similar
representations across the lifespan---may provide a simpler and more
transparent account of development.

We use a case study from word learning literature. Stager \& Werker
(1997) first showed that children's early ability to distinguish similar
sounds does not automatically translate into word learning skills.
Indeed, though infants around 14-month old can distinguish similar sound
pairs such as ``dih'' and ``bih'', they appear to fail in mapping this
pair to two different objects. Follow-up studies has focused on
proposing possible explanations for this observed gap between speech
perception and word learning (e.g., Fennell \& Waxman, 2010; Hofer \&
Levy, 2017; Rost \& McMurray, 2009; Stager \& Werker, 1997).

@@

Subsequent findings have suggested otherwise. On the one hand,
14-month-olds---who typically fail in the original tas---succeed when an
easier testing method is used, even under the same learning conditions
(Yoshida, Fennell, Swingley, \& Werker, 2009). They also succeed when
uncertainty is mitigated via disambiguating cues (e.g., Thiessen, 2007).
On the other hand, adults show patterns of learning similar to those
shown by 14-month-olds when the task is more challenging and when
similarity between words increases (Pajak, Creel, \& Levy, 2016; White,
Yee, Blumstein, \& Morgan, 2013).

@@

\begin{SCfigure}
\centering
\includegraphics[width=1.5in]{figs/model.png}
\caption{Graphical representation of our model. Circles indicate random variables (shading indicates observed variables). The squares indicate fixed model parameters.}
\label{fig:model}
\end{SCfigure}

A similar probabilistic structure was used by Lewis \& Frank (2013) to
model concept learning, and by Hofer \& Levy (2017) to model spoken word
learning. However, the first study assumed that the sounds are heard
unambiguously, and the second assumed the concepts are observed
unambiguously. In our model, we assume that both labels and concepts are
observed with a certain amount of perceptual noise, which we assume, for
simplicity, is captured by a normal distribution:

@@

The learner hears a sound \(s\) and has to decide which object \(o\)
provides an optimal match to this sound. To this end, they must compute
the probability \(P(o|s)\) for all possible objects. This probability
can be computed by summing over all possible concepts and labels:
\[P(o|s)=\sum_{C,L} P(o, C, L| s) \propto \sum_{C,L} P(o, C, L, s) \]
The joint probability \(P(o, C, L, s)\) is obtained by factoring the
Bayesian network in Figure \ref{fig:model}:
\[P(o,C,L,s) = P(s|L)P(L|C)P(C|o)P(o) \] which can be transformed using
Bayes rule into:

\[P(o,C,L,s) = P(s|L)P(L|C)P(o|C)P(C) \] Finally, assuming that the
concepts' prior probability is uniformly
distributed\footnote{This is a reasonable assumption in our particular case given the similarity of the concepts used in each naming situation in our experiment.},
we obtain the following expression, where all conditional dependencies
are now well defined:

@@

We use the model to predict performance in the word learning task
introduced by Stager \& Werker (1997), with a two-alternative forced
choice as in Yoshida et al. (2009). In this task, participants are first
exposed to the association between pairs of nonsense words (e.g.,
``lif''/``neem'') and pairs of objects. The word-object associations are
introduced sequentially. After this exposure phase, participants perform
a series of test trials. In each of these trials, one of the two sounds
is uttered (e.g., ``lif'') and participants choose the corresponding
object from the two alternatives. An overview of the task is shown in
Figure \ref{fig:task}.

We used Equation 1 to obtain the exact analytical expression for the
probability of accurate responses \(p(o_T | s)\) (target object \(o_T\)
given a sound \(s\)) in the simple case of two-alternative forced choice
in the testing phase of our experimental task:

@@

In this experiment, we tested participants in the word learning task
introduced above (Figure \ref{fig:task}). More precisely, we explored
the predictions related to both distinctiveness and precision. Sound
similarity (\(\Delta s\)) and object similarity (\(\Delta o\)) were
varied simultaneously in a within-subject design. Two age groups
(preschool children and adults) were tested on the same task to explore
whether development can be characterized with the uncertainty
parameters, \(\sigma_C\) and \(\sigma_L\). The experiment, sample size,
exclusion criteria and the model's main predictions were pre-registered.

@@

We planned to recruit a sample of N=60 children ages 4-5 years from the
Bing Nursery School on Stanford University's campus. Here we report data
from n=47 children. An additional n=28 children participated but were
removed from analyses because they were not above chance on the catch
trials due to the challenging nature of our procedure (see below). We
also planned to recruit a sample of N=60 adults on Amazon Mechanical
Turk. Data from n=10 participants were excluded due to low scores on the
catch trials (n=9 ) or because they were familiar with the non-English
sound stimuli we used in the adult experiment (n=1), yielding a final
sample of n=50.

@@

In a separate survey, \(N=20\) participants recruited on Amazon
Mechanical Turk evaluated the similarity of each sound and object pair
on a 7-point scale. We scaled responses within the range {[}0,1{]}. Data
are shown in Figure \ref{fig:stim}, for each stimulus group. These data
will be used in the models as the perceptual distance of sound pairs
(\(\Delta s\)) and object pairs (\(\Delta o\)).

@@

Each age group saw only two of the three levels of similarity described
in the previous sub-section: ``different'' vs. ``intermediate'' for
preschoolers, and ``intermediate'' vs. ``similar'' for adults. We made
this choice in light of pilot studies showing that adults were at
ceiling with ``different'' sounds/objects, and children were at chance
with the ``similar'' sounds/objects. That said, this difference in the
level of similarity is accounted for in the model through using the
appropriate perceptual distance used in each age group (Figure
\ref{fig:stim}).

@@

\caption{\label{fig:stim}Distances for both sound and object pairs from an adult norming study. Data represent Likert values normalized to [0,1] interval. Error bars represent 95\% confidence intervals.}\label{fig:stim}
\end{figure}
\end{CodeChunk}

To maximize our ability to measure subtle stimulus effects, the
experiment was a 2x2 within-subjects factorial design with four
conditions: high/low sound similarity crossed with high/low visual
object similarity. Besides the 4 conditions, we also tested participants
on a fifth catch condition which was similar in its structure to the
other ones, but was used only to select participants who were able to
follow the instructions and show minimal learning.

@@

Preschoolers were tested at the nursery school using a tablet, whereas
adults used their own computers to complete the same experiment online.
Participants were tested in a sequence of five conditions: the four
experimental conditions plus the catch condition. In each condition,
participants saw a first block of four exposure trials followed by four
testing trials, and a second block of two exposure trials (for memory
refreshment) followed by an additional four testing trials. The length
of this procedure was demanding, especially for children, but we adopted
a fully within-subjects design based on pilot testing that indicated
that precision of measurement was critical for testing our experimental
predictions.

@@

\caption{\label{fig:data_all}Accuracy of novel word recognition as as a function of the sound distance, the object distance, and the age group (preschool children vs. Adults). We show both experimental results (solid lines) and model predictions (dashed lines). Error bars represent 95\% confidence intervals.}\label{fig:all_data}
\end{figure*}
\end{CodeChunk}

@@

Experimental results are shown in Figure 6 (top). We first analyzed the
results using a mixed-effects logistic regression with sound distance,
object distance and age group as fixed effects, and with a maximal
random effects structure (allowing us to take into account the full
nested structure of our data) (Barr, Levy, Scheepers, \& Tily, 2013). We
found main effects for all the fixed effects in the regression. For the
sound distance, we obtained \(\beta =\) 0.44 (\(p\) \textless{} 0.001),
replicating previous findings. For object distance, we found \(\beta =\)
0.5 (\(p\) \textless{} 0.001), and this finding confirms the new
prediction of our model. Finally, for the age group, we obtained
\(\beta =\) 0.43 (\(p\) \textless{} 0.001), showing that performance
improves with age. In addition, we found two-way interactions between
sound distance and age (\(\beta =\) 0.17, \(p\) \textless{}0.05) and
between object distance and age (\(\beta =\) 0.18, \(p\)
\textless{}0.05).

We next fit our model (using Equation 2) to the participants' responses
in each age group. The values of \(\Delta s\) and \(\Delta o\) were set
based on data from the similarity judgment task (Figure \ref{fig:stim}).
Thus, the model has two degrees of freedom for each group, i.e.,
\(\sigma_C\) and \(\sigma_L\). Figure \ref{fig:data_all} (bottom) shows
the predictions. The model captures the qualitative patterns in both age
groups: starting from a low accuracy recognition when both the sound and
object distances are small, the model correctly predicts an increase in
accuracy when either the sound distance or the object distance
increases. Further, accuracy is correctly predicted to be maximal when
both the sound and object distances are high.

The values of the parameters were as follows. Children had a
label-specific uncertainty of \(\sigma_S =\) 0.9 {[}0.68,
1.11{]}\footnote{All uncertainty intervals in this paper represent 95\% Confidence Intervals.},
and a concept-specific uncertainty of \(\sigma_C =\) 0.29 {[}0.1,
0.49{]}. Adults had a label-specific uncertainty of \(\sigma_S =\) 0.14
{[}-0.02, 0.3{]}, and a concept-specific uncertainty of \(\sigma_C =\)
0.19 {[}0.04, 0.34{]}. As predicted, the uncertainty parameters were
larger for children than they were for adults, showing that the
probabilistic representations becomes more refined (that is, \(\sigma\)
becomes smaller) across development. The developmental effect was more
important for the label-specific uncertainty.

The models explained the majority of the variance in the participants'
mean responses (\(R^2=\) 0.86, for the combined adult and children's
data). To investigate whether the model's predictive power was due to
overfitting, we fit a simplified version with only one degree of freedom
(i.e., a single variance common to both sounds and objects). This
single-variance model explained as much variance in the mean responses
(\(R^2=\) 0.87). It also captured the main qualitative patterns (graph
not shown), suggesting that the explanatory power of the model is
largely due to its structure, rather than its degrees of freedom.

An unexpected outcome was that adult participants deviated slightly from
the model's numbers: While the model predicted accuracy to be more
sensitive to object distance when sound distance is higher (and
vice-versa), adult participants showed the opposite pattern. This
deviation is an interesting starting point for future work as it may
suggest that participants are more likely to pay attention to and
integrate additional sources of information when ambiguity is higher.

@@

This paper explored the idea that some seemingly stage-like patterns in
cognitive development can be characterized in a continuous fashion. We
used as a case study the seminal work of Stager \& Werker (1997) showing
a discrepancy between children's speech perception abilities and their
word learning skills. While much of the previous investigation of this
finding has been interested in the source of this discrepancy, here we
have explored how it could arise from continuous developmental change in
perceptual uncertainty.

@@

The model made a new prediction to which which we tested experimentally:
Learning similar words is not only modulated by the similarity of their
phonological forms, but also by the visual similarity of their semantic
referents. More generally, since visual similarity is an early
organizing feature in the semantic domain (e.g., Wojcik \& Saffran,
2013), our finding suggests that children may prioritize the acquisition
of words that are quite distant in the semantic space. This suggestion
is supported by recent findings based on the investigation of early
vocabulary growth (Engelthaler \& Hills, 2017; Sizemore, Karuza, Giusti,
\& Bassett, 2018).

One limitation of this work is that the model was fit to data from
children at a relatively older age (4-5 years old) than what is
typically studied in the literature (14-18 month-old). We selected this
age group based on two measurement issues. First, model fitting requires
several, reliably-measured data points. In our case, data collection
involved presenting participants with several trials across four
conditions in a between-subject design. It would have been challenging
to obtain these measures with infants. Second, experiments with infants
on this topic have often used habituation methods (but cf. Yoshida et
al., 2009), which require additional linking assumptions. Indeed, it is
not clear how data from these methods would directly compare to data
collected using more typical methods in older children and adults.

In sum, this paper proposes a model that accounts for the development of
an important aspect of word learning. Our account suggests that the
developmental data can be explained based on a continuous process
operating over similar representations across development, suggesting
developmental continuity. We used a case from word learning as an
example, but the same idea might apply to other aspects of cognitive
development that are typically thought of as stage-like (e.g.,
acquisition of a theory of mind). Computational models such as the one
proposed here can help us investigate the extent to which such
discontinuities emerge due to genuine qualitative changes, and the
extent to which they reflect the granularity of the researchers' own
measurement tools.
